Towards a Sound and Complete Library for Modal Logics Reasoning in Coq

### 2022-05-06 paper/paper.tex
More updates on text. Some litte progress on lean formalization. Updating gitignore

\usepackage{minted}

@@

While modal logics allow the modelling of such requirements, proving properties
about them may be a hard task without proper tool support. There are some
dedicated provers for modal logics~\cite{Gleissner17,Mora11}, but we believe
that using mature tools, like a proof assistant, are a better option, for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities that current proof assistants support allows that some trivial facts can
be checked using decision procedures.

@@

\paragraph{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The versions of
modal logics considered in this work extends traditional propositional logics
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax of modal logic formulas are defined by the following context free
grammar (where $\circ \in\{\land,\lor,\to,\leftrightarrow\}$, $\mathcal{P}$ is
the enumerable set of propositional variables and $A\in\mathcal{P}$ is arbitrary):
\[
  \begin{array}{lcl}
    \varphi & ::= & \bot\,|\,\top\,|\,A\,|\,\neg\varphi\,|\,\varphi\circ\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it express \emph{necessity}: $\Box\,\varphi$ is true at the
current world iff $\varphi$ is true at all worlds. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true at the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world.

@@

\begin{enumerate}
  \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $\mathcal{V}(A,w) = T$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall w \in W. w R y \to \mathcal{M}\,;\,y \models\varphi$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists w \in W. w R y \land \mathcal{M}\,;\,y \models\varphi$;
\end{enumerate}

We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. M\,;w\,\models \varphi$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. Finally, we say that a formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

@@

\paragraph{Syntax of modal logics formulas}
In order to implement our modal logic library, we use a \emph{deep embedding} of
formulas, i.e. we represent the syntax as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive modalFormula : Set :=
| Lit     : nat          -> modalFormula
| Neg     : modalFormula -> modalFormula
| Box     : modalFormula -> modalFormula
| Dia     : modalFormula -> modalFormula
| And     : modalFormula -> modalFormula -> modalFormula
| Or      : modalFormula -> modalFormula -> modalFormula
| Implies : modalFormula -> modalFormula -> modalFormula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$.

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by a
a type \mintinline{coq}{W : Set} to represent the set of worlds and the
accessibility relation \mintinline{coq}{R : W -> W -> Prop}.
\begin{minted}{coq}
Record Frame : Type := {
  W : Set;
  R : W -> W -> Prop;
}.
\end{minted}
A model is denoted by another record formed by a frame and a labelling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model : Type := {
  F : Frame; 
  v : nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labelling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame \mintinline{coq}{F}.

@@

>>>>

\usepackage{minted}
\usepackage{bussproof}

@@

While modal logics allow the modelling of such requirements, proving properties
about them may be a hard task without proper tool support. There are some
dedicated provers for modal logics~\cite{Gleissner17,Mora11}, but we believe
that using mature tools, like a proof assistant, are a better option, for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities that current proof assistants support, allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

\paragraph{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The versions of
modal logics considered in this work extends traditional propositional logics
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax of modal logic formulas are defined by the following context free
grammar (where $\circ \in\{\land,\lor,\to,\leftrightarrow\}$, $\mathcal{P} = \{A,B,C,...\}$ is
the enumerable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \bot\,|\,\top\,|\,A\,|\,\neg\varphi\,|\,\varphi\circ\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it express \emph{necessity}: $\Box\,\varphi$ is true at the
current world iff $\varphi$ is true at all worlds. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true at the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world.

@@

\begin{figure}[H]
\begin{enumerate}
  \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $\mathcal{V}(A,w) = T$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall w \in W. w R y \to \mathcal{M}\,;\,y \models\varphi$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists w \in W. w R y \land \mathcal{M}\,;\,y \models\varphi$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. M\,;w\,\models \varphi$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

Using the semantics, we can define the concept of \emph{entailment}: Given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi' \to \mathcal{M} \models \varphi$, for some model $\mathcal{M}$.

In this work, we consider Hilbert deductive systems for modal logics.
Essentially, the considered system adds two axioms for the modalities
to the set of axioms for classical propositional logic.
The included axioms express that $\Box$ distributes over implication and
$\Diamond$ over disjunction.
\[
  \begin{array}{ll}
    \textbf{K}           & \Box(A \to B) \to (\Box A \to \Box B)\\
    \textbf{Possibility} & \Diamond (A \lor B) \to (\Diamond A \lor \Diamond B)\\
  \end{array}
\]
Apart from the axioms, we also need the traditional \emph{modus ponens} rule
and a rule that allows to conclude formulas with necessity.

\begin{figure}[H]
\begin{tabular}{cc}
    \AxiomC{$\Gamma\vdash\varphi_1\to\varphi_2$}
    \AxiomC{$\Gamma\vdash\varphi_2$}
    \RightLabel{MP}
    \BinaryInfC{$\Gamma\vdash\varphi_2$}
    \DisplayProof
  &
    \AxiomC{$\Gamma\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\end{tabular}
\end{figure}
The set of axioms and the two previous rules forms a sound and complete Hilbert-style
deductive system for modal logics~\cite{blackburn}.

@@

\paragraph{Syntax and semantics of modal logics}
In order to implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form : Set :=
| Lit     : nat          -> form
| Neg     : form -> form
| Box     : form -> form
| Dia     : form -> form
| And     : form -> form -> form
| Or      : form -> form -> form
| Implies : form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$.

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by a
a type \mintinline{coq}{W : Set} to represent the set of worlds and the
accessibility relation is denoted by a type \mintinline{coq}{R : W -> W -> Prop}.
\begin{minted}{coq}
Record Frame : Type := {
  W : Set;
  R : W -> W -> Prop;
}.
\end{minted}
A model is denoted by another record formed by a frame and a labelling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model : Type := {
  F : Frame; 
  v : nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labelling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}. The encoding of the Kripke semantics is done by a Coq function
that which implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labelling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
and use the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model)
                           (w: W (F M))
                           (phi: form): Prop :=
  match phi with
  | Lit     x            => v M x w 
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                   formula_valuation M w' psi
  (** some code omitted* )
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{Systems for modal logic}

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of on four
possibilities in a proof: 1) the use of a hypothesis; 2) use of an axiom;
3) use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used in several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
The \mintinline{coq}{Ax} constructor deserves a bit of explanation. Function
\mintinline{coq}{instantiate} build a formula from a given axiom set for a given
system represented by a theory, which parameterizes our deduction type.
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

\paragraph{Formalizations of modal logics on proof assistants}

The formalization of epistemic modal logic in Coq was the subject of Wind's
work~\cite{dewind}.
The author made a Coq implementation of modal logic based on the S5 system,
which consists of  the axioms $K$, $T$, $4$ and $5$. The objective was the
modelling of two well known puzzles.
Wind's formalization was separated into some components. The first consists in
defining the syntax of formulas and the Kripke semantics. Next, the author uses
their definitions to give some examples of modal operators, worlds and how they
interact with each other. Using the built machinery, the author developed
introduction and elimination rules for modal logic and use them to model the
wize man's and muddy children puzzles.
